Exploring Local Explanations of Non-linear Models Using Animated Linear Projections

### 2022-01-12 _paper.rmd
chocolates case study re-written

The chocolates dataset consists of 88 observations of ten nutritional measurements determined from their labels, and labeled as being either milk or dark. Dark chocolates are typically considered healthier and often can be marketed at a higher price. It could be interesting to examine the nutritional properties of any dark chocolates that have been misclassified as milk, to determine if a manufacturer is being honest. A random forest model is used, and a chocolate labeled dark but predicted to be milk is chosen as the observation of interest (chocolate 22). It is compared with a correctly classified dark chocolate.

@@

Figure \@ref(fig:casechocolates) similarly shows that attribution-space is more separable than data-space. Interestingly, the class imbalance that we suspected was not observed; there are only six chocolates labeled as dark and predicted as milk, while eight of the inverse case. We look at one extreme instance and vary the contribution of `fiber`. We see notice two instances that have clearly well within the milk chocolate despite being observed as dark chocolate on their product label. While conjecture, there are plausible reasons that a manufacturer has incentives to cut corners and label their products different than what they are. In the penguin case we explored how the model misclassified an observation from a type one error perspective. In contrast, in the chocolates cases shows a type two error view; looking for extreme support of the models classification to challenge the original observation.

>>>>

The chocolates dataset consists of 88 observations of ten nutritional measurements determined from their labels, and labeled as being either milk or dark. Dark chocolate is considered healthier than milk. The data was collected by students during the Iowa State University  class STAT503 from nutritional information from the manufacturer's web site, and normalized to 100g equivalents. The data is available in the `cheem` package. A random forest model is used for the classification of chocolate type.

It could be interesting to examine the nutritional properties of any dark chocolates that have been misclassified as milk. A reason to do this is that a dark chocolate that is nutritionally more like milk should not be considered as a healthy alternative. It is interesting to explore which of the nutritional variables contribute most to the misclassification.

@@

This type of exploration is shown in \@ref(fig:casechocolates), where a chocolate labeled dark but predicted to be milk is chosen as the observation of interest (observation 22). It is compared with a correctly classified dark chocolate (observation 7). The PCA plot, and the SHAP PCA plots show a big difference between the two chocolate types but with confusion for a handful of observations. The misclassifications are clearer in the observed vs predicted plot, and can be seens to be mistaken in both ways: milk to dark and dark to milk.

The attribution projection for chocolate 22 suggests that Fiber, Sugars and Calories are most responsible for its incorrect prediction. The way to read this plot is to see that Fiber has a large negative value, while Sugars and Calories have reasonably large positive values. In the density plot, observations on the very left of the display would have high values of Fiber (matching the negative projection coefficient) and low values of Sugars and Calories. The opposite would be the interpretation of a point with high values in this plot. The dark chocolates (orange) are mostly on the left, and this is a reason why they are considered to be healthier: high fibre and low sugar. The density for milk chocolates is further to the right, indicating that they generally have low fibre and high sugar.

The observation of interest (dashed line) can be viewed against the comparison observation (dotted line). Now one needs to pay different attention to the parallel plot of the SHAP values, which are local to a particular observation, and the density plot which is the same projection of all observations as specifed by the SHAP values of the observation of interest.

From the parallel coordinate plot we can quickly compare the variable contributions to the two different predictions. The observation of interest differs form the comparison primarily on the Fiber variable, which suggests that this is the reason for the incorrect prediction.

From the density plot, which is the attribution projection corresponding to the observation of interest, both observations are more like dark chocolates. If we vary the contribution of Sugars, and completely remove Sugars from the projection, this is where the difference becomes apparent. When primarily Fiber is examined observation 22 looks more like a milk chocolate.

<!-- Figure \@ref(fig:casechocolates) similarly shows that attribution-space is more separable than data-space. Interestingly, the class imbalance that we suspected was not observed; there are only six chocolates labeled as dark and predicted as milk, while eight of the inverse case. We look at one extreme instance and vary the contribution of `fiber`. We see notice two instances that have clearly well within the milk chocolate despite being observed as dark chocolate on their product label. While conjecture, there are plausible reasons that a manufacturer has incentives to cut corners and label their products different than what they are. In the penguin case we explored how the model misclassified an observation from a type one error perspective. In contrast, in the chocolates cases shows a type two error view; looking for extreme support of the models classification to challenge the original observation.-->

### 2022-01-12 _paper.tex
chocolates case study re-written

The chocolates dataset consists of 88 observations of ten nutritional measurements determined from their labels, and labeled as being either milk or dark. Dark chocolates are typically considered healthier and often can be marketed at a higher price. It could be interesting to examine the nutritional properties of any dark chocolates that have been misclassified as milk, to determine if a manufacturer is being honest. A random forest model is used, and a chocolate labeled dark but predicted to be milk is chosen as the observation of interest (chocolate 22). It is compared with a correctly classified dark chocolate.

@@

Figure \ref{fig:casechocolates} similarly shows that attribution-space is more separable than data-space. Interestingly, the class imbalance that we suspected was not observed; there are only six chocolates labeled as dark and predicted as milk, while eight of the inverse case. We look at one extreme instance and vary the contribution of \texttt{fiber}. We see notice two instances that have clearly well within the milk chocolate despite being observed as dark chocolate on their product label. While conjecture, there are plausible reasons that a manufacturer has incentives to cut corners and label their products different than what they are. In the penguin case we explored how the model misclassified an observation from a type one error perspective. In contrast, in the chocolates cases shows a type two error view; looking for extreme support of the models classification to challenge the original observation.

>>>>

The chocolates dataset consists of 88 observations of ten nutritional measurements determined from their labels, and labeled as being either milk or dark. Dark chocolate is considered healthier than milk. The data was collected by students during the Iowa State University class STAT503 from nutritional information from the manufacturer's web site, and normalized to 100g equivalents. The data is available in the \texttt{cheem} package. A random forest model is used for the classification of chocolate type.

It could be interesting to examine the nutritional properties of any dark chocolates that have been misclassified as milk. A reason to do this is that a dark chocolate that is nutritionally more like milk should not be considered as a healthy alternative. It is interesting to explore which of the nutritional variables contribute most to the misclassification.

@@

This type of exploration is shown in \ref{fig:casechocolates}, where a chocolate labeled dark but predicted to be milk is chosen as the observation of interest (observation 22). It is compared with a correctly classified dark chocolate (observation 7). The PCA plot, and the SHAP PCA plots show a big difference between the two chocolate types but with confusion for a handful of observations. The misclassifications are clearer in the observed vs predicted plot, and can be seens to be mistaken in both ways: milk to dark and dark to milk.

The attribution projection for chocolate 22 suggests that Fiber, Sugars and Calories are most responsible for its incorrect prediction. The way to read this plot is to see that Fiber has a large negative value, while Sugars and Calories have reasonably large positive values. In the density plot, observations on the very left of the display would have high values of Fiber (matching the negative projection coefficient) and low values of Sugars and Calories. The opposite would be the interpretation of a point with high values in this plot. The dark chocolates (orange) are mostly on the left, and this is a reason why they are considered to be healthier: high fibre and low sugar. The density for milk chocolates is further to the right, indicating that they generally have low fibre and high sugar.

The observation of interest (dashed line) can be viewed against the comparison observation (dotted line). Now one needs to pay different attention to the parallel plot of the SHAP values, which are local to a particular observation, and the density plot which is the same projection of all observations as specifed by the SHAP values of the observation of interest.

From the parallel coordinate plot we can quickly compare the variable contributions to the two different predictions. The observation of interest differs form the comparison primarily on the Fiber variable, which suggests that this is the reason for the incorrect prediction.

From the density plot, which is the attribution projection corresponding to the observation of interest, both observations are more like dark chocolates. If we vary the contribution of Sugars, and completely remove Sugars from the projection, this is where the difference becomes apparent. When primarily Fiber is examined observation 22 looks more like a milk chocolate.
